---
title: "Group_Project"
author: "Ruolin Huang,Weifan Liu"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
---
Group members' names:
Ruolin Huang, Hangjun Liu, Weifan Liu

https://github.com/Ruolin030/BIS15L_W2026_Group10
```{r}
library("tidyverse")
library("naniar")
library("janitor")
library("ggthemes")
library("RColorBrewer")
library("paletteer")
library("ggplot2")
library("shiny")
```
1.load data.
```{r}
gestation<- read_csv("data/Danis and Rokas 2024 Gestation Length/Data_S1.csv") %>% clean_names()
```
2. Exploration of the Dataset
```{r}
glimpse(gestation)
```

```{r}
summary(gestation)
```

```{r}
dim(gestation)
```

3. Data Integrity and Quality Check
Before proceeding to analysis, we must ensure there are no hidden missing values or logical errors.
```{r}
gestation %>% 
  tabyl(order)
```

```{r}
gestation %>% 
  miss_var_summary()
```

```{r}
names(gestation)
```

```{r}
gestation %>% 
  filter(bodymass == 0 | gestationength == 0 | longevity == 0)
```

The diagnostic check confirms that the dataset is highly complete, with no missing values (NAs) or illogical zero-entries in the primary variables
#LHJ
```{r}
avg_longevity <- gestation %>%
  group_by(order) %>%
  summarise(mean_longevity = mean(longevity, na.rm = TRUE))
```

```{r}
avg_longevity %>% 
  ggplot(aes(x = order, y=mean_longevity))+
  geom_bar(stat = "identity", fill = "darkblue") +
  labs(title = "Average Longevity by Mammal Order",
       x = "order",
       y = "Average Longevity (years)")+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

3. Briefly understand the datasets.
3.1 Distrubution of order
```{r}
order_dist <- gestation %>% 
  count(order) %>% 
  mutate(order=fct_reorder(order,n))
```

```{r}
my_color <- c("#F7A6AC","#FC757B","#F97F5F","#FFA288","#FAA26F","#FDCD94","#FEE199","#FFD47D","#B0D6A9","#6AD1A3","#65BDBA","#3C9BC9","#7FBDDA","#84ADDC","#B8E5FA","#CBE4B1")
```

```{r}
order_dist %>% 
  ggplot(aes(x=order,y=n,fill=order))+
  geom_col(color="white",linewidth=0.2)+
  coord_flip()+
  geom_text(aes(label=n),hjust=-0.2,size=3)+
  scale_fill_manual(values=my_color)+
  theme_minimal()+
  theme(legend.position="none",
        panel.grid.minor=element_blank(),
        panel.grid.major.y=element_blank())+
  labs(title="Taxonomic Composition of the Mammal Dataset",
       subtitle="Analysis of 845 species organized by Order",
       x="Order",
       y="Number of Species")
```

3.2 Do larger mammals have longer pregnancies?
```{r}
gestation %>% 
  ggplot(aes(x=log10(bodymass),
             y=log10(gestationength)))+
  geom_point(aes(color=order),alpha=0.6,size=1.5)+
  geom_smooth(method="lm",color="black",linetype="dashed",se=TRUE)+
  theme_minimal()+
  scale_color_manual(values = my_color)+
  labs(title="The Allometric Relationship of Gestation Length",
       subtitle="Log-linear relationship across 845 mammalian species",
       x="Log10 Body Mass (g)",
       y="Log10 Gestation Length (days)",
       color="Order")+
  theme(legend.position="right")
```

3.3 Faceting by Order
```{r}
major_orders <- 
  gestation %>% 
  count(order) %>% 
  filter(n>=20) %>% 
  pull(order)
```

```{r}
gestation %>% 
  filter(order %in% major_orders) %>% 
  ggplot(aes(x=log10(bodymass),
             y=log10(gestationength),
             color=order))+
  geom_point(alpha=0.6,size=1)+
  facet_wrap(~order,scale="free")+
  scale_color_manual(values=my_color)+
  theme_light()+
  theme(legend.position="none",
        strip.text=element_text(face="bold",color="black"),
        strip.background=element_rect(fill="grey90"))+
  labs(title="Comparing the slopes across major mammalian groups as discussed in the paper",
       x="Log10 Body Mass(g)",
       y="log10 Gestation Length(days")
```

```{r}
gestation %>% 
  filter(order %in% major_orders) %>% 
  ggplot(aes(x=longevity,
             y=log10(gestationength),
             color=order))+
  geom_point(alpha=0.6,size=1)+
  facet_wrap(~order,scale="free")+
  scale_color_manual(values=my_color)+
  theme_light()+
  theme(legend.position="none",
        strip.text=element_text(face="bold",color="black"),
        strip.background=element_rect(fill="grey90"))+
  labs(title="Comparing the slopes across major mammalian groups as discussed in the paper",
       x="Longevity",
       y="log10 Gestation Length(days")
```


3.4 Life-history strategies （Do long-lived orders, such as primates, spend a higher percentage of their lifespan during pregnancy than those fast-dying rodents?）
```{r}
gestation_investment <- 
  gestation %>% 
  filter(order %in% major_orders) %>% 
  mutate(invest_ratio=(gestationength/360)/longevity*100)
```

```{r}
gestation_investment %>% 
  ggplot(aes(x=fct_reorder(order,invest_ratio, .desc=TRUE),
             y=invest_ratio,
             fill=order))+
  geom_boxplot(outlier.shape=NA,alpha=0.7)+
  coord_flip()+
  scale_fill_manual(values=my_color)+
  theme_minimal()+
  theme(legend.position="none")+
  labs(title="Reproductive Investment Across Mammalian Orders",
       subtitle="Percentage of lifespan spent in a single gestation period",
       x="Order",
       y="Gestation as % of Total Lifespan")
```

3.5 Body Size vs. Pregnancy's "Extreme Contrast Map"
```{r}
extreme <- gestation %>% 
  filter(!is.na(bodymass)) %>% 
  mutate(category = case_when(
    rank(bodymass) <= 10 ~ "Tiny",
    rank(-bodymass) <= 10 ~ "Giant",
    TRUE ~ NA_character_
  )) %>% 
  filter(!is.na(category))
```

```{r}
extreme %>% 
  ggplot(aes(x=fct_reorder(species,bodymass),
             y=gestationength,
             fill=category))+
  geom_col(width=0.7)+
  coord_flip()+
  scale_fill_manual(values=c("Giant"="#F7B2C7",
                             "Tiny"="#B8E5FA"))+
  theme_minimal()+
  theme(axis.text.y=element_text(face="italic",size=9),
        legend.position="top",
        panel.grid.major.y = element_blank())+
  labs(title="The Great Gestation Gap",
       subtitle="Comparing the 10 largest and 10 smallest mammals in the dataset",
       x=NULL,
       y="Gestation Length (Days)",
       fill="Group")
```

```{r}
library(shiny)

ui <- fluidPage(
  titlePanel("Mammalian Gestation Predictor"),
  
  sidebarLayout(
    sidebarPanel(
      helpText("Adjust parameters to predict gestation length."),
      
      sliderInput("mass", "Body Mass (g):", min = 2, max = 2000, value = 70, step = 10),
      sliderInput("longevity", "Lifespan (years):", min = 1, max = 200, value = 20),
      selectInput("order", "Taxonomic Order:", 
                  choices = sort(unique(gestation$order))),
      
      textOutput("equation")
    ),
    
    mainPanel(
      wellPanel(
        textOutput("prediction_text")
      ),
      
       checkboxInput("show_lm",
                "show best-fit line(lm)",
                value=FALSE),
      
      plotOutput("distPlot", width="500px", height="400px")
    )
  )
)

server <- function(input, output) {
  
  model <- lm(log(gestationength) ~ log(bodymass) + log(longevity), data = gestation)
  
  output$equation <- renderText({
    "log(gestation) = β₀ + β₁(log bodymass) + β₂(log longevity) + order"
  })
  
  output$prediction_text <- renderText({
    
    new_data <- data.frame(
      bodymass = input$mass, 
      longevity = input$longevity, 
      order = input$order
    )
    
    pred_log <- predict(model, newdata = new_data)
    pred_days <- round(exp(pred_log), 1)
    
    paste0("Predicted Gestation Length: ", pred_days, " days")
  })
  
  output$distPlot <- renderPlot({
     #current_prediction_value
    current_pred_val <- exp(predict(model, data.frame(bodymass=input$mass, # exp() gives the actual predicted gestation length in days
                                                      longevity=input$longevity, 
                                                      order=input$order)))
    
    gestation %>%
      ggplot(aes(x = bodymass, y = gestationength))+
      geom_point(alpha = 0.5, aes(color = order))+
      scale_x_log10()+
      annotate("point", x = input$mass, y = current_pred_val, 
               color = "black", size = 5, shape = 18) +
      labs(title = "Gestation vs Body Mass (Log Scale)",
           subtitle = "Black diamond indicates your current prediction",
           x = "Body Mass (g)", y = "Gestation Length (days)")+
      theme_bw()+
      if (input$show_lm)
      geom_smooth(method="lm", se=F, na.rm=T)
  else
    NULL
    
  })
}

shinyApp(ui = ui, server = server)

```

```{r}
ui <- pageWithSidebar(
  
  headerPanel("Species Finder by Traits"),
  
  sidebarPanel(
    selectInput("order",
                "Select Order",
                choices = c("All", sort(unique(gestation$order))),
                selected = "All"),
    
    sliderInput("bodymass",
                "Body Mass",
                min = min(gestation$bodymass, na.rm = TRUE),
                max = max(gestation$bodymass, na.rm = TRUE),
                value = range(gestation$bodymass, na.rm = TRUE)),
    
    sliderInput("longevity",
                "Longevity",
                min = min(gestation$longevity, na.rm = TRUE),
                max = max(gestation$longevity, na.rm = TRUE),
                value = range(gestation$longevity, na.rm = TRUE)),
    
    sliderInput("gestation",
                "Gestation Length",
                min = min(gestation$gestationength, na.rm = TRUE),
                max = max(gestation$gestationength, na.rm = TRUE),
                value = range(gestation$gestationength, na.rm = TRUE))
  ),
  
  mainPanel(
    tableOutput("species_table")
  )
)

server <- function(input, output, session) {
  
filtered_species <- reactive({
    data <- gestation
    if(input$order != "All"){data <- data %>% filter(order == input$order)}%>%
      filter(bodymass >= input$bodymass[1] & bodymass <= input$bodymass[2],
             longevity >= input$longevity[1] & longevity <= input$longevity[2],
             gestationength >= input$gestation[1] & gestationength <= input$gestation[2])})
output$species_table <- renderTable({filtered_species()})}

shinyApp(ui = ui, server = server)
```

