---
title: "Group_Project"
author: "Ruolin Huang,Weifan Liu"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
---
Group members' names:
Ruolin Huang, Hangjun Liu, Weifan Liu

https://github.com/Ruolin030/BIS15L_W2026_Group10
```{r}
library("tidyverse")
library("naniar")
library("janitor")
library("ggthemes")
library("RColorBrewer")
library("paletteer")
library("ggplot2")
library("shiny")
```
1.load data.
```{r}
gestation<- read_csv("data/Danis and Rokas 2024 Gestation Length/Data_S1.csv") %>% clean_names()
```
2. Exploration of the Dataset
```{r}
glimpse(gestation)
```

```{r}
summary(gestation)
```

```{r}
dim(gestation)
```

3. Data Integrity and Quality Check
Before proceeding to analysis, we must ensure there are no hidden missing values or logical errors.
```{r}
gestation %>% 
  tabyl(order)
```

```{r}
gestation %>% 
  miss_var_summary()
```

```{r}
names(gestation)
```

```{r}
gestation %>% 
  filter(bodymass == 0 | gestationength == 0 | longevity == 0)
```

The diagnostic check confirms that the dataset is highly complete, with no missing values (NAs) or illogical zero-entries in the primary variables
