---
title: "Group_Project"
author: "Ruolin Huang,Weifan Liu"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
---
Group members' names:
Ruolin Huang, Hangjun Liu, Weifan Liu

https://github.com/Ruolin030/BIS15L_W2026_Group10
```{r}
library("tidyverse")
library("naniar")
library("janitor")
library("ggthemes")
library("RColorBrewer")
library("paletteer")
library("ggplot2")
library("shiny")
```
1.load data.
```{r}
gestation<- read_csv("data/Danis and Rokas 2024 Gestation Length/Data_S1.csv") %>% clean_names()
```



2. Exploration of the Dataset
```{r}
glimpse(gestation)
```

```{r}
summary(gestation)
```

```{r}
dim(gestation)
```

3. Data Integrity and Quality Check
Before proceeding to analysis, we must ensure there are no hidden missing values or logical errors.
```{r}
gestation %>% 
  tabyl(order)
```

```{r}
gestation %>% 
  miss_var_summary()
```

```{r}
names(gestation)
```

```{r}
gestation %>% 
  filter(bodymass == 0 | gestationength == 0 | longevity == 0)
```

The diagnostic check confirms that the dataset is highly complete, with no missing values (NAs) or illogical zero-entries in the primary variables




```{r}
avg_longevity <- gestation %>%
  group_by(order) %>%
  summarise(mean_longevity = mean(longevity, na.rm = TRUE))
```

```{r}
avg_longevity %>% 
  ggplot(aes(x = order, y=mean_longevity))+
  geom_bar(stat = "identity", fill = "darkblue") +
  labs(title = "Average Longevity by Mammal Order",
       x = "order",
       y = "Average Longevity (years)")+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```



```{r}
ui <- pageWithSidebar(
  
  headerPanel("Species Finder by Traits"),
  
  sidebarPanel(
    selectInput("order",
                "Select Order",
                choices = c("All", sort(unique(gestation$order))),
                selected = "All"),
    
    sliderInput("bodymass",
                "Body Mass",
                min = min(gestation$bodymass, na.rm = TRUE),
                max = max(gestation$bodymass, na.rm = TRUE),
                value = range(gestation$bodymass, na.rm = TRUE)),
    
    sliderInput("longevity",
                "Longevity",
                min = min(gestation$longevity, na.rm = TRUE),
                max = max(gestation$longevity, na.rm = TRUE),
                value = range(gestation$longevity, na.rm = TRUE)),
    
    sliderInput("gestation",
                "Gestation Length",
                min = min(gestation$gestationength, na.rm = TRUE),
                max = max(gestation$gestationength, na.rm = TRUE),
                value = range(gestation$gestationength, na.rm = TRUE))
  ),
  
  mainPanel(
    tableOutput("species_table")
  )
)

server <- function(input, output, session) {
  
filtered_species <- reactive({
    data <- gestation
    if(input$order != "All"){data <- data %>% filter(order == input$order)}%>%
      filter(bodymass >= input$bodymass[1] & bodymass <= input$bodymass[2],
             longevity >= input$longevity[1] & longevity <= input$longevity[2],
             gestationength >= input$gestation[1] & gestationength <= input$gestation[2])})
output$species_table <- renderTable({filtered_species()})}

shinyApp(ui = ui, server = server)
```


```{r}
gestation %>% 
  arrange(desc(bodymass))
```
```{r}
ui <- fluidPage(
  titlePanel("Mammalian Gestation Predictor"),
  
  sidebarLayout(
    sidebarPanel(
      helpText("Adjust parameters to predict gestation length."),
      
      sliderInput("mass", "Body Mass (g):",
                  min = 2, max = 20000, value = 1000, step = 10),
      
      sliderInput("longevity", "Lifespan (years):",
                  min = 1, max = 200, value = 20),
      
      selectInput("order_select",
                  "Select Mammal Order:",
                  choices = c("All", sort(unique(gestation$order))),
                  selected = "All"),
      
      textOutput("equation")
    ),
    
    mainPanel(
      wellPanel(
        textOutput("prediction_text")
      ),
      
      checkboxInput("show_lm",
                    "Show best-fit line (lm)",
                    value = FALSE),
      
      plotOutput("distPlot", width = "500px", height = "400px")
    )
  )
)

server <- function(input, output) {
  
  model <- lm(log(gestationength) ~ log(bodymass) + log(longevity),
              data = gestation)
  
  output$equation <- renderText({
    "log(gestation) = β0 + β1 log(bodymass) + β2 log(longevity)"
  })
  
  output$prediction_text <- renderText({
    
    new_data <- data.frame(
      bodymass = input$mass,
      longevity = input$longevity
    )
    
    pred_log <- predict(model, newdata = new_data)
    pred_days <- round(exp(pred_log), 1)
    
    paste0("Predicted Gestation Length: ", pred_days, " days")
  })
  
  output$distPlot <- renderPlot({
    
    current_pred_val <- exp(
      predict(model,
              data.frame(bodymass = input$mass,
                         longevity = input$longevity))
    )
    
    filtered_data <- gestation
    
    if (input$order_select != "All") {
      filtered_data <- gestation %>%
        filter(order == input$order_select)
    }
    
    # Plot style changes depending on order selection
    if (input$order_select == "All") {
      
      p <- filtered_data %>%
        ggplot(aes(x = bodymass, y = gestationength, color = order)) +
        geom_point(alpha = 0.5)
      
    } else {
      
      p <- filtered_data %>%
        ggplot(aes(x = bodymass, y = gestationength)) +
        geom_point(color = "steelblue", size = 3) +
        geom_text(aes(label = species),
                  size = 2,
                  vjust = -0.7)
      
    }
    
    p <- p +
      scale_x_log10() +
      annotate("point",
               x = input$mass,
               y = current_pred_val,
               color = "black",
               size = 5,
               shape = 18) +
      labs(title = "Gestation vs Body Mass+lifesapn",
           subtitle = "Black diamond indicates your prediction",
           x = "Body Mass+lifesapn (log scale)",
           y = "Gestation Length") +
      theme_bw()
    
    if (input$show_lm) {
      p + geom_smooth(method = "lm", se = FALSE)
    } else {
      p
    }
    
  })
}

shinyApp(ui = ui, server = server)
```



