---
title: "Group_Project"
author: "Ruolin Huang,Weifan Liu"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
---
Group members' names:
Ruolin Huang, Hangjun Liu, Weifan Liu

https://github.com/Ruolin030/BIS15L_W2026_Group10
```{r}
library("tidyverse")
library("naniar")
library("janitor")
library("ggthemes")
library("RColorBrewer")
library("paletteer")
library("ggplot2")
library("shiny")
```
1.load data.
```{r}
gestation<- read_csv("data/Danis and Rokas 2024 Gestation Length/Data_S1.csv") %>% clean_names()
```
2. Exploration of the Dataset
```{r}
glimpse(gestation)
```

```{r}
summary(gestation)
```

```{r}
dim(gestation)
```

3. Data Integrity and Quality Check
Before proceeding to analysis, we must ensure there are no hidden missing values or logical errors.
```{r}
gestation %>% 
  tabyl(order)
```

```{r}
gestation %>% 
  miss_var_summary()
```

```{r}
names(gestation)
```

```{r}
gestation %>% 
  filter(bodymass == 0 | gestationength == 0 | longevity == 0)
```

The diagnostic check confirms that the dataset is highly complete, with no missing values (NAs) or illogical zero-entries in the primary variables
```{r}
ui <- pageWithSidebar(
  
  headerPanel("Species Finder by Traits"),
  
  sidebarPanel(
    
    sliderInput("bodymass",
                "Body Mass",
                min = min(gestation$bodymass, na.rm = TRUE),
                max = max(gestation$bodymass, na.rm = TRUE),
                value = range(gestation$bodymass, na.rm = TRUE)),
    
    sliderInput("longevity",
                "Longevity",
                min = min(gestation$longevity, na.rm = TRUE),
                max = max(gestation$longevity, na.rm = TRUE),
                value = range(gestation$longevity, na.rm = TRUE)),
    
    sliderInput("gestation",
                "Gestation Length",
                min = min(gestation$gestationength, na.rm = TRUE),
                max = max(gestation$gestationength, na.rm = TRUE),
                value = range(gestation$gestationength, na.rm = TRUE))
  ),
  
  mainPanel(
    tableOutput("species_table")
  )
)

server <- function(input, output, session) {
  
  filtered_species <- reactive({
    
    gestation %>%
      filter(
        bodymass >= input$bodymass[1] & bodymass <= input$bodymass[2],
        longevity >= input$longevity[1] & longevity <= input$longevity[2],
        gestationength >= input$gestation[1] & gestationength <= input$gestation[2]
      )
    
  })
  
  output$species_table <- renderTable({
    filtered_species()
  })
}

shinyApp(ui = ui, server = server)
```



```{r}
avg_longevity <- gestation %>%
  group_by(order) %>%
  summarise(mean_longevity = mean(longevity, na.rm = TRUE))
```

```{r}
avg_longevity %>% 
  ggplot(aes(x = order, y=mean_longevity))+
  geom_bar(stat = "identity", fill = "darkblue") +
  labs(title = "Average Longevity by Mammal Order",
       x = "order",
       y = "Average Longevity (years)")+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```